The Rockerverse: Images, Packages, and Applications for Containerization with R

### 2020-01-23 manuscript.tex
update author contributions, fix tex error, and update .tex file

% !TeX root = RJwrapper.tex
\title{The Rockerverse: Images, Packages, and Applications for Containerization
with R}
\author{by Daniel Nüst, Dom Bennett, Robrecht Cannoodt, Dav Clark, Gergely Daroczi, Dirk Eddelbuettel, Mark Edmondson, Colin Fay, Ellis Hughes, Sean Lopp, Ben Marwick, Heather Nolis, Jacqueline Nolis, Hong Ooi, Karthik Ram, Noam Ross, Lori Shepherd, Nitesh Turaga, Craig Willis, Nan Xiao}

@@

\abstract{%
The Rocker~Project provides widely-used Docker images for R across
different application scenarios. This articles surveys downstream
projects building upon Rocker and presents the current state of R
packages for managing Docker images and controlling containers. We also
look beyond Rocker to other projects connecting containerisation with R,
namely alternative suites of images. These use cases and the variety of
applications demonstrate the power of Rocker specifically and
containerisation in general. We identified common themes across this
diversity: reproducible environments, scalability and effectivity, and
portability across clouds.
}

@@

The article continues with a brief introduction of containerization
basics and the Rocker~Project, followed by descriptions of projects
developing images with R installations. Then, we present applications,
starting with the R packages specificly for interacting with Docker,
second-level packages using containers mediately or only for specific
features, up to complex use cases leveraging containers. We conclude
with a reflection on the landscape of packages and applications and
point out future directions of development.

\begin{figure}
\centering
\includegraphics{/rockerverse/rockerverse.png}
\caption{Rockerverse hex sticker}
\end{figure}

@@

The key accomplishment of Docker as an ``application'' is to make a
``bundled'' aggregation of software (the so-called ``image'') available
to any system equipped to run Docker, without requiring much else from
the host besides the actual Docker application installation. This is a
rather attractive proposition and Docker's very easy to use user
interface has lead to widespread adoption and use of Docker in a variety
of domains, e.g., cloud computing infrastructure
\citep[e.g.,][]{Bernstein2014}, data science
\citep[e.g.,][]{boettiger_introduction_2015}, and edge computing
\citep[e.g.,][]{alam_orchestration_2018}. It provided to be a natural
match for ``cloud deployment'' which runs, or at least appears to run,
``seamlessly'' without much explicit reference to the underlying
machine, architecture or operating system: containers are portable and
can be deployed with very little in terms of dependencies on the host
system---only the container runtime is required. Images are normally
build from plain text documents called \texttt{Dockerfile}. A
\texttt{Dockerfile} has a specific set of instructions to create and
document a well-defined environment, i.e., install specific software and
expose specific ports.

For statistical computing and analysis centered around R, the
Rocker~Project has provided a variety of Docker containers since its
start in 2014 \citep{RJ-2017-065}. The Rocker~Project provides several
lines of containers spanning to from building blocks with
\texttt{R-release} or \texttt{R-devel}, via containers with
\href{https://rstudio.com/products/rstudio/}{RStudio~Server} and
\href{https://rstudio.com/products/shiny/shiny-server/}{Shiny~Server},
to domain-specific containers such as
\href{https://github.com/rocker-org/geospatial}{\texttt{rocker/geospatial}}
\citep{rocker_geospatial_2019}. These containers form \emph{image
stacks}, building on top of each other for better maintainability
(i.e.~smaller \texttt{Dockerfile}s), composability, and to reduce build
time. Also of note is a series of ``versioned'' containers which match
the R release they contain with the \emph{then-current} set of packages
via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}. The
Rocker~Project's impact and importance was acknowledged by the Chan
Zuckerberg Initiative's \emph{Essential Open Source Software for
Science}, who provide findung for the projects's sustainable
maintenance, community growth, and targeting new hardware platforms
including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

@@

\hypertarget{images-for-alternative-r-distributions}{%
\subsection{Images for alternative R
distributions}\label{images-for-alternative-r-distributions}}

@@

As outlined above, R is a widely-used language with a large community.
The large number of extension packages provides access to an unrivaled
variety of established and upcoming features. Nevertheless, special use
cases and experimental projects in academia and industry exist to test
new approaches or provide features different to what \emph{GNU-R}, or
Base R, provides. GNU-R is the R distribution maintained by the
\href{https://www.r-project.org/contributors.html}{R Core Team} and
available via \href{https://cran.r-project.org/}{CRAN}. While the images
presented in this section are far from being as vetted, stable, and
widely used as any of the Rocker images, they demonstrate an important
advantage of containerisation technology, namely the abiliy to
transparently build portable stacks of open source software and make
them easily accessible to others. All these distributions of R are
published under GPL licenses. Since all of the R distributions claim
better performance as a core motivation, a comparision based on Docker
images, potentially leveraging the
\href{https://docs.docker.com/config/containers/resource_constraints/}{resource
restriction mechanisms} of Docker to level the playing field, seems
useful future work.

\textbf{Microsoft R Open} (MRO) is an R distribution formerly known as
Revolution R Open (RRO) before Revolution Analytics was acquired by
Microsoft. MRO is compatible with main R and its packages and it
``{[}..{]} includes additional capabilities for improved performance,
reproducibility, and platform support.'' \citep{microsoft_mro} Most
notably these capabilities are the MRAN repository, which is enabled by
default, and the (optional) integration with
\href{https://software.intel.com/en-us/mkl}{Intel® Math Kernel Library}
(MKL) for multi-threading in linear algebra operations
\citep{microsoft_multithread}. MRO does not provide official Docker
images, but a set of community-maintained \texttt{Dockerfile}s and
Docker images are provided by the poject \texttt{mro-docker}
(\url{https://github.com/nuest/mro-docker}), e.g.~the image
\href{https://hub.docker.com/repository/docker/nuest/mro/}{\texttt{nuest/mro}}.
The images are inspired by the Rocker images and can be used much in the
same fashion, effectively a drop-in replacement allowing users to
quickly evaluate if the benefits of MRO + Intel® MKL apply to their use
case. Version-tagged images are provided for the latest bugfix release
of recent R versions. Extended license information about MKL is printed
at every startup.

\textbf{Renjin} is an interpreter for the R language running on top of
the \href{https://en.wikipedia.org/wiki/Java_virtual_machine}{Java
Virtual Machine} (JVM) providing full two-way access between Java and R
code \citep{wikipedia_renjin_2018}. It was developed to combine the
benefits of R, such scripting and extension packages, with the JVM's
advantages in the areas of security, cross-platform availability, and
established position in enterprise settings. R extension packages need
to be specially compiled and are distributed via the Java package
manager \href{https://en.wikipedia.org/wiki/Apache_Maven}{Apache Maven},
cf. \url{http://packages.renjin.org/packages} for available packages.
Packages are loaded on demand, i.e., at the first call to
\texttt{library()}. Not all R packages, especially one linking to binary
libraries, are available, e.g., \texttt{rgdal}
\footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
There are no offical Docker images for Renjin, but community-maintained
images for selected releases only are available under
\texttt{nuest/renjin} on Docker~Hub and GitHub at
\url{https://hub.docker.com/r/nuest/renjin} and
\url{https://github.com/nuest/renjin-docker} respectively. These images
expose the command line interface of Renjin in a similar fashion as
Rocker images and allow an easy evaluation of Renjin's suitability, but
are not intended for production use.

@@

\textbf{FastR} (\url{https://github.com/oracle/fastr}) is ``A
high-performance implementation of the R programming language, built on
GraalVM'' \citep{oracle_labs_oraclefastr_2020}. It is developed by
Oracle, connects R to the GraalVM ecosystem
\citep{wikipedia_graalvm_2019}, and also claims superior performance but
also targets full compatibility with base R
\citep{oracle_labs_oraclefastr_2020}. There are no offical Docker images
provided, but indepedently maintained experimental images and
\texttt{Dockerfile}s are provided by
\href{https://github.com/nuest/fastr-docker}{nuest/fastr-docker}, e.g.
\href{https://cloud.docker.com/repository/docker/nuest/fastr/}{\texttt{nuest/fastr}
on Docker Hub}.

@@

\emph{Bioconductor} (\url{https://bioconductor.org/}) is an open source,
open development project for the analysis and comprehension of genomic
data \citep{gentleman_bioconductor_2004}. The project consists of 1741 R
software packages as of August 15th 2019, as well as packages containing
annotation or experiment data. \emph{Bioconductor} has a semiannual
release cycle, each release is associated with a particular version of
R. Docker images allow availability of current and past versions of
\emph{Bioconductor} for convenience and reproducibility.
\emph{Bioconductor}'s `base' docker images are built on top of
\texttt{rocker/r-ver} and \texttt{rocker/rstudio} images.
\emph{Bioconductor} installs packages based on the R version, and
therefore uses \texttt{rocker/rstudio} and \texttt{rocker/r-ver} version
tagging. \emph{Bioconductor} selects the desired version of R from
Rocker, adds the BiocManager CRAN package for installing appropriate
versions of \emph{Bioconductor} packages, and creates a
\emph{Bioconductor} docker image with an informative tag
(R\_version\_Bioc\_version). The images are summarized on the
\emph{Bioconductor} web site (see
\url{https://bioconductor.org/help/docker/}), created with
\texttt{Dockerfile}s maintained on GitHub
(\url{https://github.com/Bioconductor/bioconductor_docker}), and
\href{https://hub.docker.com/u/bioconductor}{distributed through
DockerHub}. Past and current combinations of R and \emph{Bioconductor}
are therefore accessible via a specific image tag. A strength of this
approach is that the responsibility for complex software configuration
(including customized development) is shifted from the user to the
experienced \emph{Bioconductor} core team. Images may be valuable in
teaching, where participants pull pre-built images to avoid complicated
configuration of their own computing environemnts.

\emph{Bioconductor} has several images in addition to `base', specific
to various areas of research. The `core' image installs the most
commonly used \emph{Bioconductor} packages. \emph{Bioconductor} images
for proteomics, metabolomics, and flow cytometry are community
maintained. All community maintained images build on top of the
\emph{Bioconductor} base image and therefore indirectly the Rocker
images. To simplify building and maintaining \emph{Bioconductor} images,
we use a Ruby templating engine. A recent audit of the
\emph{Bioconductor} \texttt{Dockerfile}s, following best practices from
the Docker website, led to a reduction in the size and number of layers.
The most important insights involve the
\href{https://docs.docker.com/storage/storagedriver/overlayfs-driver/\#how-container-reads-and-writes-work-with-overlay-or-overlay2}{`union'
file system used by Docker}. In this file system, once a layer (e.g.,
\texttt{RUN} statement) writes to a file path, the file path is never
altered. A subsequent layer that might appear to remove or overwrite the
path actually masks, rather than alters, the original. It is therefore
important to clean up (e.g., cache removal) within each layer, and to
avoid re-installing existing dependencies.

A recent innovation is to produce a \texttt{bioconductor\_full} image to
emulate the \emph{Bioconductor} nightly Linux build machine. The image
contains the \emph{system dependencies} needed to install and check
almost all (1730 of 1741) \emph{Bioconductor} software packages. Users
no longer have to manage complciated system dependencies. The image is
configured so that \texttt{.libPaths()} has
\texttt{/usr/local/lib/R/host-site-library} as the first location. Users
mounting a location on the host file system to this location then
persist installed packages across docker sessions or updates. Many
\emph{R} users pursue flexible work flows tailored to particular
analysis needs, rather than standardized work flows. The
\texttt{bioconductor\_full} image is well-suited to this pattern.
\texttt{bioconductor\_full} provides developers with a test environment
like \emph{Bioconductor}'s build system.

@@

\textbf{rocker-win} (\url{https://github.com/nuest/rocker-win}) is a
proof of concept for running R in Windows-based containers
\citep{nust_rocker-win_2019}. It provides selected R versions using
three different base images using \texttt{microsoft/windowsservercore}
and \texttt{mcr.microsoft.com/windows/servercore}. These base images
match different Windows versions: Windows Server 2019, Windows Server
2016, and Windows Server, version 1803. The images are built using two
CI services,
\href{https://docs.travis-ci.com/user/reference/windows/}{Travis CI} and
\href{https://www.appveyor.com/docs/windows-images-software/}{Appveyor},
which provide the different Windows versions and therefore support
different base images, and published automatically
\href{https://hub.docker.com/r/nuest/rocker-win}{on Docker~Hub}. The
images are built from manually maintained seperate \texttt{Dockerfile}s,
are tagged with both Windows-variant and R version, e.g.,
\texttt{nuest/rocker-win:ltsc2019-3.6.2} or
\texttt{nuest/rocker-win:1803-latest}, and run the \texttt{R.exe}
process by default. All these images can be run on Docker for Desktop on
Windows 10. New R versions are only added on demand. The explorative
rocker-win project demonstrates the variety and pontentially confusing
complexity of Windows on Docker, e.g., because of the different base
images, but also provides a starting point for any Windows-constrained
user to leverage containerisation, including examples for apps using
\textbackslash{}pkg\{plumber, \pkg{Shiny}, or R packages with system
dependencies such as \pkg{sf}. The latter is quite comfortable and fast
actually, compared to installation from source on Linux, as R ships
pre-compiled binaries for Windows. However, compared to the various
Linux images, the footprint of Windows images is quite large.

For Windows, just as originally for Linux, the cloud use cases drive the
development and Docker and Microsoft collaborate
closely\footnote{\href{https://www.docker.com/partners/microsoft}{https://www.docker.com/partners/microsoft}}.
In most cases, individual developers or researchers worrying about
reproducibility, will prefer the more widely and host independent
Linux-based containers. Naturally, a Windows license is required for the
host machine, and the licensing impacts the potnetial to build images in
clouds or redistribute exported images as archive files. But the
developments in the \texttt{rocker-win} prototype show that containers
can also be used for R workflows depending on Windows-only tools, for
leveraging an existing Windows Server-based infrastructure where
policies can otherwise not be met, or for streamlining interactions with
system operating staff that runs Windows-based containers themselves.

\hypertarget{non-debian-linux-images}{%
\subsection{Non-Debian Linux images}\label{non-debian-linux-images}}

@@

\label{rhub} The \textbf{R-hub} project provides \emph{``a collection of
services to help R package development''}, with the package builder as
the most prominent one\citep{r-hub_docs_2019}. The builder allows R
package developers to check their R package on different platforms and R
versions using a web form or the package \pkg{rhub}
\citep{csardi_rhub_2019}. The builder uses Docker containers to conduct
these checks, taking advantages of their well-defined environments and
sandboxing. The \texttt{Dockerfile}s and images are published on
Docker~Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}}
and
GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}}
respectively. The images comprise release and development builds of R
running in the base images of the
\href{https://www.debian.org/}{Debian},
\href{https://ubuntu.com/}{Ubuntu},
\href{https://getfedora.org/}{Fedora}, and
\href{https://centos.org/}{CentOS} linux distributions (Arch Linux is
under
development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}},
see \texttt{rhub::local\_check\_linux\_images()} for a full list), and
are intended for debugging (see Section~\nameref{development}) not for
reproducible workflows. \emph{The rationale} of not re-using the Rocker
image stack is the high number of operatings systems and configurations,
which can be more uniformly maintained in a seperate suite of images.
Also, a common structure across the respective image stacks of the
operating systems, e.g., for \texttt{debian}, adding GCC in
\texttt{debian-gcc} and then R's development version in
\texttt{debian-gcc-devel}. The platform also has a specific user
configuration differing from the setup covering typical applications of
Rocker images, and the images include considerably more software, such
as a fairly complete LaTeX installation, to provide an experience closer
to CRAN.

@@

There are also independent projects installing R or RStudio on
\href{https://www.alpinelinux.org/}{\textbf{Alpine Linux}}, but not
beyond the proof of concept
stage\footnote{\href{https://gitlab.com/artemklevtsov/r-alpine}{https://gitlab.com/artemklevtsov/r-alpine} or \href{https://www.github.com/velaco/alpine-r}{https://www.github.com/velaco/alpine-r} provide different use case images but are not recently maintained (development of the latter documented in \citet{ratesic_building_2018}); \href{https://github.com/cmplopes/alpine-r}{https://github.com/cmplopes/alpine-r} has only base R images, is relatively up-to-date but sparsely documented; \href{https://github.com/CenterForStatistics-UGent/mountainr}{https://github.com/CenterForStatistics-UGent/mountainr} is a ghost project with only a single short `Dockerfile`, yet it shows the easy installation of the latest R from Alpine sources}.
In general, Alpine-based Docker images are often chosen for minimalistic
and thereby small and secure containers, e.g., in hardware with limited
storage. R-hub's proof of concept
\href{https://github.com/r-hub/r-minimal}{\texttt{r-minimal}} takes this
to the extreme with compressed image size of 20MB and bespoke install
scripts, e.g., to cleverly remove compilers after package installation.
For R though, this advantage quickly deteriorates if more features are
needed, e.g.~if adding compilers, packages, or especially if installing
powerful data science and communication tools, like RStudio or LaTeX.
Furthermore the distribution uses the
\href{https://www.musl-libc.org/}{\texttt{musl} C library} instead of
the \texttt{glibc} used in Debian/Ubuntu, for which more tooling and
experiences
exist\footnote{Cf. \href{https://github.com/rocker-org/rocker/issues/231}{https://github.com/rocker-org/rocker/issues/231}}.
In fact, using Debian's `slim' variant and removing capabilities, one
could probalby achieve similarly small image sizes on a more common
stack.

@@

Each of these packages has particular advantages as they provide
function wrappers for interacting with the Docker CLI at different
stages of a container's life cycle. Examples of such interactions are
installing the Docker software, creating \texttt{Dockerfile}s
(\pkg{dockerfiler}), \pkg{containerit}), building images and launching a
containers (\pkg{stevedore}, \pkg{docker}) on a local machine or on the
cloud. As such, the choice of which package is most useful depends on
the use-case at hand, but also the users level of expertise.

>>>>

% !TeX root = RJwrapper.tex
\title{The Rockerverse: Images, Packages, and Applications for Containerization
with R}
\author{by Daniel Nüst, Dom Bennett, Robrecht Cannoodt, Dav Clark, Gergely Daroczi, Dirk Eddelbuettel, Mark Edmondson, Colin Fay, Ellis Hughes, Sean Lopp, Ben Marwick, Heather Nolis, Jacqueline Nolis, Hong Ooi, Karthik Ram, Noam Ross, Lori Shepherd, Nitesh Turaga, Craig Willis, Nan Xiao, Charlotte Van Petegem}

@@

\abstract{%
The Rocker~Project provides widely-used Docker images for R across
different application scenarios. This articles surveys downstream
projects building upon Rocker and presents the current state of R
packages for managing Docker images and controlling containers. We also
look beyond Rocker to other projects connecting containerisation with R,
namely alternative suites of images. These use cases and the variety of
applications demonstrate the power of Rocker specifically and
containerisation in general. We identified common themes across this
diversity: reproducible environments, scalability and efficiency, and
portability across clouds.
}

@@

The article continues with a brief introduction of containerization
basics and the Rocker~Project, followed by descriptions of projects
developing images with R installations. Then, we present applications,
starting with the R packages specificly for interacting with Docker,
second-level packages using containers indirectly or only for specific
features, up to complex use cases leveraging containers. We conclude
with a reflection on the landscape of packages and applications and
point out future directions of development.

@@

The key accomplishment of Docker as an ``application'' is to make a
``bundled'' aggregation of software (the so-called ``image'') available
to any system equipped to run Docker, without requiring much else from
the host besides the actual Docker application installation. This is a
rather attractive proposition and Docker's very easy to use user
interface has lead to widespread adoption and use of Docker in a variety
of domains, e.g., cloud computing infrastructure
\citep[e.g.,][]{Bernstein2014}, data science
\citep[e.g.,][]{boettiger_introduction_2015}, and edge computing
\citep[e.g.,][]{alam_orchestration_2018}. It provided to be a natural
match for ``cloud deployment'' which runs, or at least appears to run,
``seamlessly'' without much explicit reference to the underlying
machine, architecture or operating system: containers are portable and
can be deployed with very little in terms of dependencies on the host
system---only the container runtime is required. Images are normally
built from plain text documents called \texttt{Dockerfile}. A
\texttt{Dockerfile} has a specific set of instructions to create and
document a well-defined environment, i.e., install specific software and
expose specific ports.

For statistical computing and analysis centered around R, the
Rocker~Project has provided a variety of Docker containers since its
start in 2014 \citep{RJ-2017-065}. The Rocker~Project provides several
lines of containers spanning to from building blocks with
\texttt{R-release} or \texttt{R-devel}, via containers with
\href{https://rstudio.com/products/rstudio/}{RStudio~Server} and
\href{https://rstudio.com/products/shiny/shiny-server/}{Shiny~Server},
to domain-specific containers such as
\href{https://github.com/rocker-org/geospatial}{\texttt{rocker/geospatial}}
\citep{rocker_geospatial_2019}. These containers form \emph{image
stacks}, building on top of each other for better maintainability (i.e.,
smaller \texttt{Dockerfile}s), composability, and to reduce build time.
Also of note is a series of ``versioned'' containers which match the R
release they contain with the \emph{then-current} set of packages via
the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}. The
Rocker~Project's impact and importance was acknowledged by the Chan
Zuckerberg Initiative's \emph{Essential Open Source Software for
Science}, who provide findung for the projects's sustainable
maintenance, community growth, and targeting new hardware platforms
including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

@@

\hypertarget{images-for-alternative-r-engines}{%
\subsection{Images for alternative R
engines}\label{images-for-alternative-r-engines}}

@@

As outlined above, R is a widely-used language with a large community.
The large number of extension packages provides access to an unrivaled
variety of established and upcoming features. Nevertheless, special use
cases and experimental projects in academia and industry exist to test
new approaches, or provide features different to what the R
implementation maintained by the
\href{https://www.r-project.org/contributors.html}{R Core Team} and
available via \href{https://cran.r-project.org/}{CRAN} provides. While
the images presented in this section are far from being as vetted,
stable, and widely used as any of the Rocker images, they demonstrate an
important advantage of containerisation technology, namely the abiliy to
transparently build portable stacks of open source software and make
them easily accessible to others. All these alternative implementations
of R are published under GPL licenses. Since all implementations of R
claim better performance as a core motivation, a comparision based on
Docker images, potentially leveraging the
\href{https://docs.docker.com/config/containers/resource_constraints/}{resource
restriction mechanisms} of Docker to level the playing field, seems
useful future work.

\textbf{Microsoft R Open} (MRO) is an R distribution formerly known as
Revolution R Open (RRO) before Revolution Analytics was acquired by
Microsoft. MRO is compatible with main R and its packages and it
``{[}..{]} includes additional capabilities for improved performance,
reproducibility, and platform support.'' \citep{microsoft_mro} Most
notably these capabilities are the MRAN repository, which is enabled by
default, and some bundled packages such as \CRANpkg{checkpoint} and
specific R packages maintained by
Microsoft\footnote{See "Bundled Packages" on \href{https://mran.microsoft.com/rro/installed}{https://mran.microsoft.com/rro/installed}.}.
Originally the optional integration with
\href{https://software.intel.com/en-us/mkl}{Intel® Math Kernel Library}
(MKL) for multi-threading in linear algebra operations
\citep{microsoft_multithread} was a distinguishing feature before it was
made available publicly, and can be used in base R, too
\citep{eddelbuettel_thinking_2018}.

MRO does not provide official Docker images, but a set of
community-maintained \texttt{Dockerfile}s and Docker images are provided
by the poject \texttt{mro-docker}
(\url{https://github.com/nuest/mro-docker}), e.g., the image
\href{https://hub.docker.com/repository/docker/nuest/mro/}{\texttt{nuest/mro}}.
The images are inspired by the Rocker images and can be used much in the
same fashion, effectively a drop-in replacement allowing users to
quickly evaluate if the benefits of MRO + Intel® MKL apply to their use
case. Version-tagged images are provided for the latest bugfix release
of recent R versions. Extended license information about MKL is printed
at every startup.

\textbf{Renjin} is an interpreter for the R language running on top of
the \href{https://en.wikipedia.org/wiki/Java_virtual_machine}{Java
Virtual Machine} (JVM) providing full two-way access between Java and R
code \citep{wikipedia_renjin_2018}. It was developed to combine the
benefits of R, such scripting and extension packages, with the JVM's
advantages in the areas of security, cross-platform availability and
integration into enterprise platforms. R extension packages need to be
specially compiled and are distributed via the Java package manager
\href{https://en.wikipedia.org/wiki/Apache_Maven}{Apache Maven},
cf.~\url{http://packages.renjin.org/packages} for available packages.
Packages are loaded on demand, i.e., at the first call to
\texttt{library()}. Not all R packages, especially one linking to binary
libraries, are available, e.g., \texttt{rgdal}
\footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
There are no offical Docker images for Renjin, but community-maintained
images for selected releases only are available under
\texttt{nuest/renjin} on Docker~Hub and GitHub at
\url{https://hub.docker.com/r/nuest/renjin} and
\url{https://github.com/nuest/renjin-docker} respectively. These images
expose the command line interface of Renjin in a similar fashion as
Rocker images and allow an easy evaluation of Renjin's suitability, but
are not intended for production use.

@@

\textbf{FastR} (\url{https://github.com/oracle/fastr}) is ``A
high-performance implementation of the R programming language, built on
GraalVM'' \citep{oracle_labs_oraclefastr_2020}. It is developed by
Oracle, connects R to the GraalVM ecosystem
\citep{wikipedia_graalvm_2019}, and also claims superior performance but
also targets full compatibility with base R
\citep{oracle_labs_oraclefastr_2020}. There are no offical Docker images
provided, but indepedently maintained experimental images and
\texttt{Dockerfile}s are provided by
\href{https://github.com/nuest/fastr-docker}{nuest/fastr-docker}, e.g.,
\href{https://cloud.docker.com/repository/docker/nuest/fastr/}{\texttt{nuest/fastr}
on Docker Hub}.

@@

\emph{Bioconductor} (\url{https://bioconductor.org/}) is an open source,
open development project for the analysis and comprehension of genomic
data \citep{gentleman_bioconductor_2004}. The project consists of 1823 R
software packages as of October 30th 2019, as well as packages
containing annotation or experiment data. \emph{Bioconductor} has a
semi-annual release cycle, each release is associated with a particular
version of R, and Docker images are provided for current and past
versions of \emph{Bioconductor} for convenience and reproducibility. All
images, included are described on the \emph{Bioconductor} web site (see
\url{https://bioconductor.org/help/docker/}), created with
\texttt{Dockerfile}s maintained on GitHub, and distributed through
Docker~Hub\footnote{See \href{https://github.com/Bioconductor/bioconductor_docker}{https://github.com/Bioconductor/bioconductor\_docker} and \href{https://hub.docker.com/u/bioconductor}{https://hub.docker.com/u/bioconductor} respectively.}.
\emph{Bioconductor}'s `base' Docker images are built on top of the
\texttt{rocker/rstudio} image. \emph{Bioconductor} installs packages
based on the R version in combination with the Bioconductor version, and
therefore uses Bioconductor version tagging \texttt{devel} and
\texttt{RELEASE\_X\_Y}, e.g., \texttt{RELEASE\_3\_10}. Past and current
combinations of R and \emph{Bioconductor} will therefore be accessible
via a specific image tags. The \emph{Bioconductor} \texttt{Dockerfile}
selects the desired version of R from Rocker, adds required system
dependencies, and uses the \CRANpkg{BiocManager} package for installing
appropriate versions of \emph{Bioconductor} packages. A strength of this
approach is that the responsibility for complex software configuration
(including customized development) is shifted from the user to the
experienced \emph{Bioconductor} core team. A recent audit of the
\emph{Bioconductor} image stack \texttt{Dockerfile} led to the
deprecation of several community-maintained images, because the numerous
specific images became too hard to understand, complex to maintain, and
cumbersome to extent. As part of the simplification, a recent innovation
is to produce a \texttt{bioconductor\_docker:devel} image to emulate the
\emph{Bioconductor} nightly Linux build machine as closely as possible.
This image contains the build system environment variables and the
\emph{system dependencies} needed to install and check almost all (1813
out of 1823) \emph{Bioconductor} software packages and saves users and
package developers from managing these themselves. Furthermore the image
is configured so that \texttt{.libPaths()} has
\texttt{/usr/local/lib/R/host-site-library} as the first location. Users
mounting a location on the host file system to this location can persist
installed packages across Docker sessions or image updates. Many R users
pursue flexible work flows tailored to particular analysis needs, rather
than standardized work flows. The new \texttt{bioconductor\_docker}
image is well-suited for this pattern, while
\texttt{bioconductor\_docker:devel} provides developers with a test
environment close to \emph{Bioconductor}'s build system.

@@

\textbf{rocker-win} (\url{https://github.com/nuest/rocker-win}) is a
proof of concept for running R in Windows-based containers
\citep{nust_rocker-win_2019}. It provides selected R versions using
three different base images using \texttt{microsoft/windowsservercore}
and \texttt{mcr.microsoft.com/windows/servercore}. These base images
match different Windows versions: Windows Server 2019, Windows Server
2016, and Windows Server, version 1803. The images are built using two
CI services,
\href{https://docs.travis-ci.com/user/reference/windows/}{Travis CI} and
\href{https://www.appveyor.com/docs/windows-images-software/}{Appveyor},
which provide the different Windows versions and therefore support
different base images, and published automatically
\href{https://hub.docker.com/r/nuest/rocker-win}{on Docker~Hub}. The
images are built from manually maintained seperate \texttt{Dockerfile}s,
are tagged with both Windows-variant and R version, e.g.,
\texttt{nuest/rocker-win:ltsc2019-3.6.2} or
\texttt{nuest/rocker-win:1803-latest}, and run the \texttt{R.exe}
process by default. All these images can be run on Docker for Desktop on
Windows 10. New R versions are only added on demand. The explorative
rocker-win project demonstrates the variety of Windows on Docker, which
is potentially confusing, e.g., because of the different base images.
The proof of concept provides a starting point for any
Windows-constrained user to leverage containerisation, including
examples for apps using \CRANpkg{plumber}, \CRANpkg{Shiny}, or R
packages with system dependencies such as \CRANpkg{sf}. The latter is
quite comfortable and fast actually, compared to the default
installation from source on Linux, because CRAN ships pre-compiled
binaries for Windows. However, compared to the various Linux images, the
footprint of Windows images is quite large.

For Windows, just as originally for Linux, the cloud use cases drive the
development and Docker and Microsoft collaborate
closely\footnote{\href{https://www.docker.com/partners/microsoft}{https://www.docker.com/partners/microsoft}}.
In most cases, individual developers or researchers worrying about
reproducibility, will prefer the more widely and host independent
Linux-based containers. Naturally, a Windows license is required for the
host machine, and the licensing impacts the potential to build images in
clouds or redistribute exported images as archive files. But the
developments in the \texttt{rocker-win} prototype show that containers
can also be used for R workflows depending on Windows-only tools, for
leveraging an existing Windows Server-based infrastructure where
policies can otherwise not be met, or for streamlining interactions with
system operating staff that runs Windows-based containers themselves.

\hypertarget{non-rocker-linux-images}{%
\subsection{Non-Rocker Linux images}\label{non-rocker-linux-images}}

@@

\label{rhub} The \textbf{R-hub} project provides \emph{``a collection of
services to help R package development''}, with the package builder as
the most prominent one\citep{r-hub_docs_2019}. The builder allows R
package developers to check their R package on different platforms and R
versions using a web form or the package \CRANpkg{rhub}
\citep{csardi_rhub_2019}. The builder uses Docker containers to conduct
these checks, taking advantages of their well-defined environments and
sandboxing. The \texttt{Dockerfile}s and images are published on
Docker~Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}}
and
GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}}
respectively. The images comprise release and development builds of R
running in the base images of the
\href{https://www.debian.org/}{Debian},
\href{https://ubuntu.com/}{Ubuntu},
\href{https://getfedora.org/}{Fedora}, and
\href{https://centos.org/}{CentOS} linux distributions (Arch Linux is
under
development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}},
see \texttt{rhub::local\_check\_linux\_images()} for a full list), and
are intended for debugging (see Section~\nameref{development}) not for
reproducible workflows. \emph{The rationale} of not re-using the Rocker
image stack is the high number of operatings systems and configurations,
which can be more uniformly maintained in a seperate suite of images.
Also, a common structure across the respective image stacks of the
operating systems, e.g., for \texttt{debian}, adding GCC in
\texttt{debian-gcc} and then R's development version in
\texttt{debian-gcc-devel}. The platform also has a specific user
configuration differing from the setup covering typical applications of
Rocker images, and the images include considerably more software, such
as a fairly complete LaTeX installation, to provide an experience closer
to CRAN.

@@

There are also independent projects installing R or RStudio on
\href{https://www.alpinelinux.org/}{\textbf{Alpine Linux}}, but not
beyond the proof of concept
stage\footnote{\href{https://gitlab.com/artemklevtsov/r-alpine}{https://gitlab.com/artemklevtsov/r-alpine} or \href{https://www.github.com/velaco/alpine-r}{https://www.github.com/velaco/alpine-r} provide different use case images but are not recently maintained (development of the latter documented in \citet{ratesic_building_2018}); \href{https://github.com/cmplopes/alpine-r}{https://github.com/cmplopes/alpine-r} has only base R images, is relatively up-to-date but sparsely documented; \href{https://github.com/CenterForStatistics-UGent/mountainr}{https://github.com/CenterForStatistics-UGent/mountainr} is a ghost project with only a single short `Dockerfile`, yet it shows the easy installation of the latest R from Alpine sources}.
In general, Alpine-based Docker images are often chosen for minimalistic
and thereby small and secure containers, e.g., in hardware with limited
storage. R-hub's proof of concept
\href{https://github.com/r-hub/r-minimal}{\texttt{r-minimal}} takes this
to the extreme with compressed image size of 20MB and bespoke install
scripts, e.g., to cleverly remove compilers after package installation.
For R though, this advantage quickly deteriorates if more features are
needed, e.g., if adding compilers, packages, or especially if installing
powerful data science and communication tools, like RStudio or LaTeX.
Furthermore the distribution uses the
\href{https://www.musl-libc.org/}{\texttt{musl} C library} instead of
the \texttt{glibc} used in Debian/Ubuntu, for which more tooling and
experiences
exist\footnote{Cf. \href{https://github.com/rocker-org/rocker/issues/231}{https://github.com/rocker-org/rocker/issues/231}}.
In fact, using Debian's `slim' variant and removing capabilities, one
could probalby achieve similarly small image sizes on a more common
stack.

@@

Each of these packages has particular advantages as they provide
function wrappers for interacting with the Docker CLI at different
stages of a container's life cycle. Examples of such interactions are
installing the Docker software, creating \texttt{Dockerfile}s
(\CRANpkg{dockerfiler}, \pkg{containerit}), building images and
launching a containers (\CRANpkg{stevedore}, \pkg{docker}) on a local
machine or on the cloud. As such, the choice of which package is most
useful depends on the use-case at hand, but also the users level of
expertise.
